PTFO 8-8695: Two Stars, Two Signals, No Planet

### 2020-06-16 paper/ms.tex
minor
made discussion better

\explain{Note: added citation to Bodman et al 2017 in this paragraph.}
These light curves have been phenomenologically classified as
``persistent flux dips'' or ``transient flux dips'', based on whether
their depths and durations show variability over the 90-day K2
campaigns \citep{stauffer_orbiting_2017}.  In the terminology of
\citet{stauffer_orbiting_2017}, these objects are morphologically
distinct from ``scallop shell'' light curves, and are present in stars
at more advanced evolutionary disk stages than the ``dipper'' stars
\citep{ansdell_young_2016,bodman_dippers_2017,cody_manyfaceted_2018}.
The persistent and transient flux dip stars all show angular dips that
are cannot be explained as the effects of starspots.  These stars
typically have the following things in common:
\begin{enumerate}[topsep=0.5ex,itemsep=-0.5ex,partopsep=1ex,parsep=1.5ex]
  \item They are weak-lined T Tauri stars.
  \item The spectral type is M2 to M5 ({\it e.g.},
    \citealt{rebull_usco_2018},~Figure~20).
  \item \replaced{The age is typically $\lesssim$ 100$\,$Myr.}{The
    ages typically range between 5 and 100$\,$Myr.}\footnote{ At
    present, the oldest observed ``scallops'' are in the Pleaides
    \citep{rebull_rotation_2016}. One of these, EPIC 211013604, might
    meet the ``persistent dip'' classification.  If so, it is \deleted{also} the
    oldest known. \added{The lower limit comes from the lack of detectable
   WISE infrared excesses.}} 
  \item The light curves show shallow, angular dips, \deleted{usually}
    superposed on large-amplitude smooth variability \added{at or near
    1:1 synchronicity}. The \replaced{latter}{smooth variability} is
    interpreted as stellar rotation.
  \item The rotation is rapid, with a period that is usually between
    0.5 and 2.0 days.
  \item There is rarely any infrared excess that is detectable in the
    WISE data (never any W4 detection; only a few W3 detections).
  \item They sometimes show multiple dips per cycle.
  \item The dip depths, durations, and phases can vary over just a few
    cycles ({\it e.g.}, EPIC 204143627).
  \item The dip depths can change after flares.
  \item They are rare at a population level, with an occurrence rate
  \replaced{of $\lesssim 1\%$}{between 0.1--1\%}
   of young M2-M5 stars \citep{rebull_usco_2018}.
\end{enumerate}
The 10.76$\,$hr signal in \ptfo\ meets all of these criteria.  This is
the first connection of \ptfo\ with this class of objects, likely
because the TESS data enabled us to put the dips in the context of the
asymmetric out-of-dip modulation.

@@

\added{However, the association between transient flux dips and
binarity is not iron-clad.  Firstly, multiple photometric periods do
not necessarily imply binarity. For instance, circumstellar material
could exist around single stars inside, at, or wide of co-rotation.
Additionally, } low-mass pre-main-sequence stars \deleted{however}
have been shown to \added{have} companion fractions up to twice as
high in dispersed clusters such as Upper~Sco and Taurus
\citep{kraus_mapping_2008,kraus_mapping_2011}.  A high-resolution
imaging survey would be interesting, to determine whether the
transient flux dip stars truly have distinct population-level binarity
properties relative to other young low-mass stars.

@@

The key question is what causes the transient dips. This is an
unsolved problem not only for \ptfo\ but also for the emerging class
of similar young and rapidly rotating M-dwarfs.  Many possible
explanations were discussed by \citet{rebull_rotation_2016},
\citet{david_transient_2017}, \citet{stauffer_orbiting_2017}, and
\citet{zhan_complex_2019}.  Among the {\it disfavored} explanations
are that the dips are caused by
{\it (i)} eclipsing binaries;
{\it (ii)} ``dipper''-flavor Class-I or Class-II disks;
{\it (iii)} eclipses of prominences;
{\it (iv)} high-latitude accretion hotspots;
{\it (v)} high-latitude starspots;
or {\it (vi)} dust clouds of \replaced{plausible
composition}{particular compositions
\citep[][Figure~10]{zhan_complex_2019}}.  We also view the possibility
of {\it (vii)} tidally disrupted planetary or cometary material to be
implausible, given the synchronicity between dip and rotation periods
seen across many systems.

The explanations that \replaced{are not yet}{have not yet been} ruled
out include {\it (i)} transiting clumps of \added{dust or }gas at the
Keplerian co-rotation radius; {\it (ii)} transits of enshrouded
protoplanets; {\it (iii)} occultations of \replaced{starspots}{hot and
cold spots on the stellar surface} by an optically thick disk.  The
first and last explanations have added appeal because they are
flexible enough to explain not only the transient and persistent-dip
M-dwarfs, but also the ``scallop shell'' M-dwarfs
\citep{stauffer_orbiting_2017}.  \replaced{Despite this appeal, the
possibility of distinct mechanisms explaining these distinct
variability classes remains open.}{ In the spot occultation scenario
\citep{zhan_complex_2019}, the ``blobs'' that provide the necessary
brightness contrast are on the surface of the star, rather than at the
co-rotation radius. The observing geometry would require an inner ring
to be projected into a narrow band in front of the stellar disk, in
order to produce variations with sufficiently short timescales.  The
outer disk would also need to either not be present, or to be
misaligned enough and insufficiently flared, that it did not fall
along the line of sight.  This geometric requirement seems too
contrived to occur in even $\lesssim$1\% of young M dwarf systems. }

\replaced{The evolution of \ptfo\ over}{Observations of \ptfo\ from}
the past decade could offer additional hints.  \deleted{Specifically,
}\ptfo's transition between having none, one, and multiple dips per
cycle seems important \added{(Figure~\ref{fig:o_minus_c})}.
\deleted{The dips would then need to be caused by material that was
somehow disrupted from the planet, but somehow remained co-orbital for
an extended duration. This seems implausible.} \added{ The H$\alpha$
excess observed by \citet{johnskrull_h_2016} also seems like an
important piece of evidence.  The model with circumstellar optically
thick hydrogen gas near the co-rotation radius readily explains this
excess.  }

\added{Out of the available explanations, we therefore consider the
idea of a clumpy torus of optically thick dust or gas near the
co-rotation radius to be most plausible.  There is theoretical support
for the idea that accretion disks subject to strong magnetic fields
could enter stalled states and become trapped near co-rotation
\citep{dangelo_accretion_2012}.  The concept of an ``evaporating dust
rim'' within which all dust evaporates
\citep{millan-gabet_circumstellar_2007,dullemond_inner_2010} will
likely also be an important first-order consideration in physical models of
this scenario.  The additional question of whether the opacity in fact
comes from dust or partially-ionized gas ({\it e.g.}, via gas
continuum, atomic lines, or perhaps TiO molecular opacity) also
remains a topic for future exploration.  }

>>>>

\explain{Note: added citation to Bodman et al 2017 in this paragraph.}
These light curves have been phenomenologically classified as
``persistent flux dips'' or ``transient flux dips'', based on whether
their depths and durations show variability over the 90-day K2
campaigns \citep{stauffer_orbiting_2017}.  In the terminology of
\citet{stauffer_orbiting_2017}, these objects are morphologically
distinct from ``scallop shell'' light curves, and are present in stars
at more advanced evolutionary disk stages than the ``dipper'' stars
\citep{ansdell_young_2016,bodman_dippers_2017,cody_manyfaceted_2018}.
The persistent and transient flux dip stars all show angular dips that
are cannot be explained as the effects of starspots.  These stars
typically have the following things in common:
\begin{enumerate}[topsep=0.5ex,itemsep=-0.5ex,partopsep=1ex,parsep=1.5ex]
  \item They are weak-lined T Tauri stars\added{ (under the
    definition of {\it e.g.}, \citealt{fang_star_2009})}.
  \item The spectral type is M2 to M5 ({\it e.g.},
    \citealt{rebull_usco_2018},~Figure~20).
  \item The age is typically $\lesssim$ 100$\,$Myr.\footnote{ At
    present, the oldest observed ``scallops'' are in the Pleaides
    \citep{rebull_rotation_2016}. One of these, EPIC 211013604, might
    meet the ``persistent dip'' classification.  If so, it is \deleted{also} the
    oldest known. } 
  \item The light curves show shallow, angular dips, \deleted{usually}
    superposed on large-amplitude smooth variability \added{at or near
    1:1 synchronicity}. The \replaced{latter}{smooth variability} is
    interpreted as stellar rotation.
  \item The rotation is rapid, with a period that is usually between
    0.5 and 2.0 days.
  \item There is rarely any infrared excess that is detectable in the
    WISE data (never any W4 detection; only a few W3 detections).
  \item They sometimes show multiple dips per cycle.
  \item The dip depths, durations, and phases can vary over just a few
    cycles ({\it e.g.}, EPIC 204143627).
  \item The dip depths can change after flares.
  \item They are rare at a population level, with an occurrence rate
  \replaced{of $\lesssim 1\%$}{between 0.1--1\%}
   of young M2-M5 stars \citep{rebull_usco_2018}.
\end{enumerate}
The 10.76$\,$hr signal in \ptfo\ meets all of these criteria.
\replaced{This is the first connection of \ptfo\ with this class of
objects, likely because the TESS data enabled us to put the dips in
the context of the asymmetric out-of-dip modulation.} {While this was
in fact mentioned by \citet{stauffer_orbiting_2017}, the
asymmetric out-of-dip modulation shown by the TESS data solidifies the
connection.}

@@

\added{However, the association between transient flux dips and
binarity is not iron-clad.  Firstly, multiple photometric periods do
not necessarily imply binarity. For instance, circumstellar material
could exist around single stars inside, at, or wide of co-rotation.
If it did, it could eclipse with a period different from the stellar
rotation.
Additionally, } low-mass pre-main-sequence stars \deleted{however}
have been shown to \added{have} companion fractions up to twice as
high in dispersed clusters such as Upper~Sco and Taurus
\citep{kraus_mapping_2008,kraus_mapping_2011}.  A high-resolution
imaging survey would be interesting, to determine whether the
transient flux dip stars truly have distinct population-level binarity
properties relative to other young low-mass stars.

@@

\explain{Note :Removed ``high-latitude accretion hotspots'' from the
list in this paragraph}
The key question is what causes the transient dips. This is an
unsolved problem not only for \ptfo\ but also for the emerging class
of similar young and rapidly rotating M-dwarfs.  Many possible
explanations were discussed by \citet{rebull_rotation_2016},
\citet{david_transient_2017}, \citet{stauffer_orbiting_2017}, and
\citet{zhan_complex_2019}.  Among the {\it disfavored} explanations
are that the dips are caused by
{\it (i)} eclipsing binaries;
{\it (ii)} ``dipper''-flavor Class-I or Class-II disks;
{\it (iii)} eclipses of prominences;
{\it (iv)} high-latitude starspots;
or {\it (v)} dust clouds of \replaced{plausible
composition}{particular compositions
\citep[][Figure~10]{zhan_complex_2019}}.  We also view the possibility
of {\it (vi)} tidally disrupted planetary or cometary material to be
implausible, given the synchronicity between dip and rotation periods
seen across many systems.  \added{Finally, {\it (vii)} transits of an
enshrouded protoplanet can be added to the list of disfavored
explanations, based on the chromaticity, asymmetric phase curve, and
light-curve changes seen in \ptfo.}

\explain{Note: ``High-latitude accretion hotspots'' was moved to this
paragraph.}
The explanations that we view as not yet ruled out include {\it A)}
transiting clumps of \added{dust or }gas at the Keplerian co-rotation
radius; {\it B)} occultations of \replaced{starspots}{hotspots on the
stellar surface} by an optically thick disk \citep{zhan_complex_2019};
and {\it C)} high-latitude accretion hotspots passing behind the star
\citep{yu_tests_2015}.  \replaced{The evolution of \ptfo\
over}{Observations of \ptfo\ from} the past decade may offer some
hints.  \deleted{Specifically, }\ptfo's transition between having
none, one, and multiple dips per cycle seems important
\added{(Figure~\ref{fig:o_minus_c})}.  \deleted{The dips would then
need to be caused by material that was somehow disrupted from the
planet, but somehow remained co-orbital for an extended duration. This
seems implausible.} \added{ The H$\alpha$ excess observed by
\citet{johnskrull_h_2016} also seems like an important piece of
evidence.  }

\explain{Note: This paragraph is newly added, to support our view of the
high-latitude accretion hotspot being plausible for \ptfo.}
The high-latitude accretion hotspot scenario seems qualitatively
capable of explaining most of the characteristics of \ptfo, including
the fact that its H$\alpha$ equivalent width of 11.4$\,$\AA\
\citep{van_eyken_ptf_2012} places it near the ``classical/weak'' T
Tauri division \citep{briceno_cidaII_2019}, and the fact that its
excess H$\alpha$ component varies in velocity between $-200$ and
$+200$ km$\,$s$^{-1}$ \citep{johnskrull_h_2016}.  The lack of a
NIR-excess could be problematic, but does not eliminate the
possibility that a tenuous amount of dust or gas remains near the
star, and is able to accrete onto it.  We view the main problem with
this explanation to be the likelihood that it explains the population
of transient flux dips.  RIK-210 in particular has dips too deep
($\sim$20\%) and too asymmetric to be explained by occultations of a
hot spot \citep{david_transient_2017}.

\explain{Note: This paragraph has been expanded to clarify the
possibility of a hotspot being occulted by a disk.}
Explanations {\it A} and {\it B} (circumstellar material at the
co-rotation radius, and occultations of spots by a disk) have the
appeal that they are flexible enough to explain not only the transient
and persistent-dip stars, but also the ``scallop shells''
\citep{stauffer_orbiting_2017}.  \replaced{Despite this appeal, the
possibility of distinct mechanisms explaining these distinct
variability classes remains open.}{ In the spot occultation scenario
\citep{zhan_complex_2019}, the ``blobs'' that provide the necessary
brightness contrast are on the surface of the star, rather than at the
co-rotation radius. The observing geometry would require an inner ring
to be projected into a narrow band in front of the stellar disk, in
order to produce variations with sufficiently short timescales.  The
outer disk would also need to either not be present, or to be
misaligned enough and insufficiently flared, that it did not fall
along the line of sight.  This geometric requirement seems too
contrived to occur in even $\lesssim$1\% of young M dwarf systems. }

\added{Out of the available explanations, we therefore consider the
clumpy torus of optically thick dust or gas near the co-rotation
radius to be most plausible.  The partially ionized circumstellar
hydrogen would readily explain the excess H$\alpha$ emission, and
changes in the clumping geometry would lead to changes in the observed
eclipse shapes.  The long-term stability, punctuated by changes in dip
phases, might be understood as an accretion disk subject to strong
magnetic fields going through ``stalled'' states in which material is
trapped near co-rotation \citep{dangelo_accretion_2012}.  Physical
models of this scenario may need to consider such states, in addition
to dissociation of the dust population
\citep{millan-gabet_circumstellar_2007,dullemond_inner_2010}.
Finally, the question of whether the opacity in fact comes from dust
or partially-ionized gas ({\it e.g.}, via gas continuum, atomic lines,
or perhaps TiO molecular opacity) remains an additional issue for
future exploration.  }
